### 2012-09-22 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
finised initial parallel mc content

\section{Monte Carlo Synthetic-Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution that was not prohibited severely by the quality of the
initial guess for the system \citep{evans_monte_2009}. Their approach
was instead to use residual Monte Carlo as a synthetic acceleration
for a stationary method. To derive this method, we begin by splitting
the operator in Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly.
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therfore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
    \label{eq:mcsa}
  \end{gather}
\end{subequations}
where the adjoint Monte Carlo method is used to generate the solution
correction from the residual. Using Monte Carlo in this way achieves
the same effect as Halton's method, decoupling its convergence rate
from the overall convergence rate of the method. Here, the approximate
Monte Carlo solution is not driven to a particular convergence as it
merely supplies a correction for the initial guess generated by
Richardson's iteration. Rather, only a set number of histories are
required using the adjoint method to generate the correction.

@@

\subsubsection{Load Balancing Concerns}
\label{subsubsec:mc_load_balancing}
Although domain decomposition was shown to be efficient in a load
balanced situation \citep{siegel_analysis_2012}, careful consideration
must be made for situations where this is not the case. Given the
stochastic nature of the problem and lack of a globally homogoneous
domain, parallel Monte Carlo simulations are inherently load
imbalanced. Procassini and others worked to alleviate some of the load
imbalances that are generated by both particle and spatial parallelism
and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to compute
require more work). In this variation, domains that require more work
will be replicated more frequently at reduced particle counts in each
replication. Consider the example in
Figure~\ref{fig:procassini_example} adapted from Procassini's work.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Example illustrating how domain decomposition can
      create load balance issues in Monte Carlo.}  \textit{A domain is
      decomposed into 4 zones on 8 processors with a point source in
      the lower left zone. As the particles diffuse from the source in
      the random walk sequence as shown in the top row, their tracks
      populate the entire domain. As given in the bottom row, as the
      global percentage of particles increases in a zone, that zone's
      replication count is increased.}}
  \label{fig:procassini_example}
\end{figure}
In this example, a geometry is decomposed into 4 domains on 8
processors with a point source in the bottom left domain. To begin,
because the point source is concentrated in one domain, that domain is
replicated 5 times such the amount of work it has to do per processor
is roughly balanced with the others. As the particles begin to diffuse
away from the point source, the amount of replication is adjusted to
maintain load balance. Near the end of the simulation, the diffusion
of particles is enough that all domains have equal replication.  By
doing this, load balance is improved as each domain has approximately
equal work although each domain may represent a different spatial
location and have a differing number of histories to
compute. Furthermore, Procassini and colleagues noted that as the
simulation progressed and particles were transported throughout the
domain, the amount of replication for each domain would vary as
particle histories began to diffuse, causing some regions to have
higher work loads and some to have smaller work loads than the initial
conditions. Compared to Wagner's work where the fission source was
distributed relatively evenly throughout the domain, fixed source
problems (and especially those that have a point-like source) like
those presented in Procassini's work will be more prone to changing
load balance requirements.

@@

Their solutions to these problems are elegant in that they require a
minimal amount of modification to be applied to the Monte Carlo
algorithm. To solve the first issue, in order to ensure each particle
maintains an invariant random number sequence that determines its
behavior regardless of domain decomposition, each particle carries
with it a random number seed that describes its current state. These
seeds are derived from the actual geometric location of the particle
such that it is decomposition invariant. Non-commutative floating
point operations are solved by instead mapping floating point values
to 64-bit integer values for which additions will always be
commutative. Once the operations are complete, these integers are
mapped back to floating point values.

\subsection{Parallel Adjoint Method}
\label{subsec:parallel_adjoint}
We can take much of what was learned from the survey of parllel Monte
Carlo methods for radiation transport and directly apply it to a
parallel formulation of our stochastic linear solvers.

\subsection{Parallel MCSA}
\label{subsec:parallel_mcsa}
It should be noted here that a lot of the parallel matrix/vector
operations discussed in the parallel projection methods section also
apply here. Three of the four MCSA steps are in fact these operations
with convergence checks also using a norm reduction.

>>>>

\section{Monte Carlo Synthetic-Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution that was not prohibited severely by the quality of the
initial guess for the system \citep{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \citep{evans_monte_2012}, achieving identical
numerical results as convential Krylov solvers. Their approach was
instead to use residual Monte Carlo as a synthetic acceleration for a
stationary method. To derive this method, we begin by splitting the
operator in Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly.
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therfore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the adjoint Monte Carlo method is used to generate the solution
correction from the residual. Using Monte Carlo in this way achieves
the same effect as Halton's method, decoupling its convergence rate
from the overall convergence rate of the method. Here, the approximate
Monte Carlo solution is not driven to a particular convergence as it
merely supplies a correction for the initial guess generated by
Richardson's iteration. Rather, only a set number of histories are
required using the adjoint method to generate the correction.

@@

\subsubsection{Load Balancing Concerns}
\label{subsubsec:mc_load_balancing}
Although domain decomposition was shown to be efficient in a load
balanced situation \citep{siegel_analysis_2012}, careful consideration
must be made for situations where this is not the case. Given the
stochastic nature of the problem and lack of a globally homogoneous
domain, parallel Monte Carlo simulations are inherently load
imbalanced. Procassini and others worked to alleviate some of the load
imbalances that are generated by both particle and spatial parallelism
and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to compute
require more work). In this variation, domains that require more work
will be replicated more frequently at reduced particle counts in each
replication. Furthermore, Procassini and colleagues noted that as the
simulation progressed and particles were transported throughout the
domain, the amount of replication for each domain would vary as
particle histories began to diffuse, causing some regions to have
higher work loads and some to have smaller work loads than the initial
conditions. Consider the example in
Figure~\ref{fig:procassini_example} adapted from Procassini's work.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Example illustrating how domain decomposition can
      create load balance issues in Monte Carlo.}  \textit{A domain is
      decomposed into 4 zones on 8 processors with a point source in
      the lower left zone. As the particles diffuse from the source in
      the random walk sequence as shown in the top row, their tracks
      populate the entire domain. As given in the bottom row, as the
      global percentage of particles increases in a zone, that zone's
      replication count is increased.}}
  \label{fig:procassini_example}
\end{figure}
In this example, a geometry is decomposed into 4 domains on 8
processors with a point source in the bottom left domain. To begin,
because the point source is concentrated in one domain, that domain is
replicated 5 times such the amount of work it has to do per processor
is roughly balanced with the others. As the particles begin to diffuse
away from the point source, the amount of replication is adjusted to
maintain load balance. Near the end of the simulation, the diffusion
of particles is enough that all domains have equal replication.  By
doing this, load balance is improved as each domain has approximately
equal work although each domain may represent a different spatial
location and have a differing number of histories to
compute. Compared to Wagner's work where the fission source was
distributed relatively evenly throughout the domain, fixed source
problems (and especially those that have a point-like source) like
those presented in Procassini's work will be more prone to changing
load balance requirements.

@@

Their solutions to these problems are elegant in that they require a
minimal amount of modification to be applied to the Monte Carlo
algorithm. To solve the first issue, in order to ensure each particle
maintains an invariant random number sequence that determines its
behavior regardless of domain decomposition, each particle is a random
number seed that describes its current state upon entering the domain
of a new processor. These seeds are derived from the actual geometric
location of the particle such that it is decomposition
invariant. Non-commutative floating point operations are solved by
instead mapping floating point values to 64-bit integer values for
which additions will always be commutative. Once the operations are
complete, these integers are mapped back to floating point values.

\subsection{Parallel Adjoint Method}
\label{subsec:parallel_adjoint}
We can take much of what was learned from the survey of parallel Monte
Carlo methods for radiation transport and directly apply it to a
parallel formulation of our stochastic linear solvers. Direct analogs
can be derived from these works by noting that the primary difference
between solving a linear system with Monte Carlo methods and fixed
source Monte Carlo transport problems is the content of Markov chains
that are generated. These transitions represeneted by these chains are
bound by probabilities and weights and are initiated by the sampling
of a source. In the context of transport problems, those transitions
represent events such as pariticle scattering and absorbition with
probabilities that are determined by physical data in the form of
cross sections. For stochastic matrix inversion, those transitions
represent moving between the equations of the linear system (and
therefore the physical domain which they represent) and their
probabilites are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains.

Therefore, parallel methods for Monte Carlo radiation transport and
abstract those that apply to matrix inversion methods as an initial
means of developing a parallel Neumann-Ulam-type solver. Based on the
results observed in the last decade of parallel Monte Carlo
development, we can generate many practical questions that this type
of work could answer. Given a decomposed domain, it is clear that
domain overlap significantly reduces the amount of communication
required between adjacent domains as histories move to states that are
not owned by the local processor. How much domain overlap is suitable
for matrix inversion problems? Are we memory limited by large problems
such that only so much overlap is feasible? Are there metrics related
to the properties of the matrix such that we can provide guidelines
for selecting the amount of overlap required? Furthermore, as Wagner
and colleagues observed \citep{wagner_hybrid_2010}, with marginal
domain overlap for reactor problems, the percentage of histories
leaving the local domain can easily be reduced to less than 1\%. Evans
and colleagues in their initial MCSA development typically used only
50 particle histories in order to compute a correction in an MCSA
iteration \citep{evans_monte_2012}. Per the Central Limit Theorem,
that correlates to a statistical uncertainty of 14.1\% in the
correction yet as compared to convential Krylov solvers, MCSA achieved
the identical numerical results. If good convergence and numerically
accurate solutions can be achieved with such a large uncertainty in
the correction computation, then perhaps with enough domain overlap
the minimal amount of histories that do transition to non-local states
can be ignored and thus eliminate all communication in the parallel
Monte Carlo adjoint solver and create an embarassingly parallel
method. It will be important to determine if the expectation value
bias in the MCSA correction generated by this approximation will
remain within the bounds of the already high statistical uncertainty
for unbiased estimates, thus providing numerically equivalent results.

From a domain replication perspective, this may be difficult to
achieve with a production scale linear solver. Typically, memory is at
a premimium and therefore the more distinct domains available in the
decomposition, the spatially finer and/or the numerically more
accurate the discretization that can be implemented. How much
replication is possible for large problems? How does replication
facilitate parallel performance when coupled with domain overlap? How
can we measure how much domain overlap is feasible? Replicating
domains may therefore run into memory limitations for exceptionally
large problems such that the operator, solution vector, and source
vector must be copied in their entirety multiple times. From a
resiliency standpoint, such an operation will be required, and
therefore its performance and memory implications on conventional
problems must be analyzed.

\subsection{Parallel MCSA}
\label{subsec:parallel_mcsa}
With a parallel adjoint Neumann-Ulam solver implementation, the
parallel implementation of the MCSA method will be trivial. Recall the
MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
vector operations as utilized in conventional Krylov methods. We
utilize these here for the parallel MCSA implementation. In the first
step, a parallel matrix-vector multiply is used to apply the split
operator to the previous iterate's solution. A parallel vector update
is then performed with the source vector to arrive at the initial
iteration guess. In the next step, the residual is computed by the
same operations where now the operator is applied to the solution
guess with a parallel matrix-vector multiply and then a parallel
vector update with the source vector is performed. Once the correction
is computed with a parallel adjoint Neumann-Ulam solve, this
correction is applied to the guess with a parallel vector update to
get the new iteration solution. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector. For
this implementation, all of the issues that will be potentially
generated by the parallel adjoint solver implementation will manifest
themselves here as the quality of the correction will be of intense
study.
